---
title: "Part I: Exploratory Data Analysis, Linear Regression, ANOVA"
output:
  pdf_document: default
  html_document: default
---

### Load Packages 

First, we must load the packages that will be used in the first part of this workshop.


```{r, message=FALSE, warning=FALSE }
library(pastecs, quietly = TRUE)
library(lm.beta,  quietly = TRUE)
library(lmtest,  quietly = TRUE)
library(foreign,  quietly = TRUE)
library(lattice,  quietly = TRUE)
library(lme4,  quietly = TRUE)
library(nlme,  quietly = TRUE)
library(survival,  quietly = TRUE)
library(dplyr,  quietly = TRUE)
library(ggfortify,  quietly = TRUE)
library(survminer,  quietly = TRUE)
library(rms,  quietly = TRUE)
```

#Exploratory Data Analysis

## Basic Statistical Analysis 
### Data set description

In this section, we will be using the iris data set. This data set contains measurement data of the flower of certain plant species. The data set has five variables:

- *Sepal.Length* - measurements of Sepal length
- *Sepal.Width* - measurements of Sepal width
- *Petal.Length* - measurements of Petal length
- *Petal.Width* - measurements of Petal width
- *Species* - species of the plant

```{r}
data <- iris
```

### Univariate descriptive statistics

The function, `stat.desc`, can be used to do a statistical analysis of data. It returns the mean, median, maximum, minimum, etc of a data set.

```{r}
stat.desc(data$Sepal.Width)

```
### Descriptive statistics by groups 

Using `tapply`, we compute the same descriptive statistics above but grouped species with the same sepal width. `tapply` takes

- *first argument*: the input data to which we will apply the statistical function 
- *second argument*: the grouping data which tells the statistical function how to group the input data
- *third argument*: the statistical function.

We will be consider the statistical functions: `mean`,`sd` and `length`.

```{r}
mean <- tapply(data$Sepal.Length, data$Sepal.Width, mean)
standard_deviation <- tapply(data$Sepal.Length, data$Sepal.Width, sd)
number_of_observations <- tapply(data$Sepal.Length, data$Sepal.Width, length)
round(cbind(mean, standard_deviation, number_of_observations), digits = 6)
```

## Basic Data Visualization


It is important get an idea of the "structure" of your data. To do this, we can use histograms. We use the `hist` function to plot a distribution of sepal width.


```{r}
hist(data$Sepal.Width,
     xlab='Sepal Width',ylab='Count',
     main= 'Histogram plot of sepal length vs sepal width')
```


We use the `plot` function to create a scatter plot of sepal length vs sepal width.

```{r}
plot(data$Sepal.Length, data$Sepal.Width,
     xlab='Sepal length', ylab='Sepal Width',
     main= 'Scatter plot of sepal length vs sepal width')
```

### Stacking plots (without ggplot2)

We also differentiate the scatter plot above by species. We plot each species separately. First, we call `plot` to create a canvas with an set of points associated with Setosa. If we were to call `plot` again, this would clear the previous plot. Rather, we call `points` to add scatter plots to existing plot. `points` does not clear the previous plot.
 
```{r}
plot(data[data$Species == "setosa", ]$Sepal.Length, 
     data[data$Species == "setosa", ]$Sepal.Width, 
     xlab = 'Sepal Length',  ylab= 'Sepal Width',
     xlim = range(as.matrix(data$Sepal.Length)), 
     ylim = range(as.matrix(data$Sepal.Width)),
     main= 'Scatter plot of sepal length vs sepal width')

points(data[data$Species == "versicolor", ]$Sepal.Length,
       data[data$Species == "versicolor", ]$Sepal.Width,
       col = 'yellow')

points(data[data$Species == "virginica", ]$Sepal.Length,
       data[data$Species == "virginica", ]$Sepal.Width, col = 'blue')
```

### Scatter plot matrix

It is quite cumbersome to call to `plot` multiple times to create scatter plots for various pairs of explanatory variables. 

There exists a convience function, `pairs`, that will create a matrix of scatter plots for all possible explanatory variable combination.

We give the `pairs` the first four columns of `data`. It will create a scatter plot matrix for sepal length, sepal width, petal length and petal width.

```{r}
pairs(data[,c(1:4)])
```

### Correlation matrix

From the scatter plot matrix above, we can see the qualitative correlation patterns between explanatory variables. We can also calculate these correlations explicitly and as a matrix using the `cor` function.

```{r}
cor(data[,c(1:4)])
```

# Linear regression

Given a response variable, $y$, explanatory variables, $X_i$, and assuming that

$$ y_i = \beta_0 +\beta_1 X_{1i} + \beta_2 X_{2i} + \cdots + \beta_n X_{ni} +\varepsilon$$

where $\varepsilon$ is a noise term, linear regression attempts the find the coefficients $\beta_i$ that makes $\varepsilon$ as small as possible.

The formula above assumes that response variable is a linear function explanatory variables. The noise term is added to account for the fact that most data is noisy and will not perfectly fit its 'true' function. 

Linear regression also assumes that the noise is normally distributed with zero mean. 

If these two assumptions are voilated then 
  $$r_i = y_i - \beta_0 +\beta_1 X_{1i} + \beta_2 X_{2i} + \cdots + \beta_n X_{ni}$$
  
  and $\sum_{i=1}^m r_i^2$ is generally a large number. $\sum_{i=1}^m r_i^2$ is called the residual standard error.
  
  


## Linear regression syntax

To regress the response variable, `Sepal.Width`, with explanatory variables, `Sepal.Length`, `Species`, `Petal.Length` and `Petal.Width`, we use the `lm` function.

The first argument is `lm` is the formula. The name of the column of the response variable is written first. It is followed by a tilde, `~`. After the tilde, we write names of the explanatory variable each separated by a `+`. `1` can also be added to the formula to represent a constant term.

### General Syntax: Constant Term

First, I use the formula `Sepal.Width ~ 1`. This formula is equivalent to

$$\text{Sepal.Width} = \beta_0 + \varepsilon.$$

Note that constant term is simply the mean response variable.

```{r}
ols <- lm(Sepal.Width ~ 1, data = data)
summary(ols)
print(mean(data$Sepal.Width))
```

```{r}
plot(data[data$Species == "setosa", ]$Sepal.Length, 
     data[data$Species == "setosa", ]$Sepal.Width, 
     xlab = 'Sepal Width',  ylab= 'Sepal Length',
     xlim = range(as.matrix(data$Sepal.Length)), 
     ylim = range(as.matrix(data$Sepal.Width)),
     main= 'Scatter plot of sepal length vs sepal width')

coefs = coef(ols)
abline(coefs[1],0)
```


### General Syntax: Explanatory Variable and Constant Term

I use then formula `Sepal.Width ~ 1 + Sepal.Length`. This formula is equivalent to

$$\text{Sepal.Width} = \beta_0 + \beta_1\text{Sepal.Length}+ \varepsilon$$
```{r}
ols <- lm(Sepal.Width ~ 1 + Sepal.Length, data = data)
summary(ols)
```

```{r}
plot(data[data$Species == "setosa", ]$Sepal.Length, 
     data[data$Species == "setosa", ]$Sepal.Width, 
     xlab = 'Sepal Width',  ylab= 'Sepal Length',
     xlim = range(as.matrix(data$Sepal.Length)), 
     ylim = range(as.matrix(data$Sepal.Width)),
     main= 'Scatter plot of sepal length vs sepal width')

coefs = coef(ols)
abline(coefs[1],coefs[2])
```

### General Syntax: Factors

I use then formula `Sepal.Width ~ 1 + Sepal.Length + as.factor(Species)`. This formula is equivalent to

$$\text{Sepal.Width} = \beta_0 + \beta_1\text{Sepal.Length} + \beta_2 I(\text{Species} = \text{Veriscolor}) + \beta_3 I(\text{Species} = \text{Virginica}) + \varepsilon$$

```{r}
ols <- lm(Sepal.Width ~ 1 + Sepal.Length + as.factor(Species), data = data)
summary(ols)
```

```{r}
plot(data[data$Species == "setosa", ]$Sepal.Length, 
     data[data$Species == "setosa", ]$Sepal.Width, 
     xlab = 'Sepal Width',  ylab= 'Sepal Length',
     xlim = range(as.matrix(data$Sepal.Length)), 
     ylim = range(as.matrix(data$Sepal.Width)),
     main= 'Scatter plot of sepal length vs sepal width')

points(data[data$Species == "versicolor", ]$Sepal.Length,
       data[data$Species == "versicolor", ]$Sepal.Width,
       col = 'yellow')

points(data[data$Species == "virginica", ]$Sepal.Length,
       data[data$Species == "virginica", ]$Sepal.Width, col = 'blue')
coefs = coef(ols)
abline(coefs[1],coefs[2])
abline(coefs[3] + coefs[1],coefs[2],col='yellow')
abline(coefs[4] +coefs[1],coefs[2],col='blue')
```

### Advanced Syntax: Nonlinear Regression

The `lm` function can be extended to nonlinear functions. For example, it possible include a quadratic term in our model.

$$\text{Sepal.Width} = \beta_1\text{Sepal.Length} + \beta_2\text{Sepal.Length}^2  + \varepsilon$$

To add a quadratic term to the model, add `I(Sepal.Length^2)` to the left side of the tilde, `~`.


```{r}
ols_quadratic <- lm(Sepal.Width ~ Sepal.Length + I(Sepal.Length^2), data = data)
summary(ols_quadratic)
```

It is also possible to include higher order nonlinear terms, such as cubic, quintic, etc.

### Advanced Syntax: Interaction term

It is also possible to include model interaction between explanatory variables. 

$$\text{Sepal.Width} = \beta_1\text{Sepal.Length} + \beta_2\text{Petal.Length} + \beta_3 \text{Sepal.Length}\times\text{Petal.Length} + \varepsilon$$
$\text{Sepal.Length}\times\text{Petal.Length}$ models simple interaction between Sepal.Length and Petal.Length.

```{r}
ols_interaction <- lm(Sepal.Width ~ Sepal.Length + Petal.Length + Sepal.Length*Petal.Length, data = data)
summary(ols_interaction)
```

Note that using the formula `Sepal.Width ~ Sepal.Length*Petal.Length` produces the same result as `Sepal.Length + Petal.Length + Sepal.Length*Petal.Length`.

```{r}
ols_interaction <- lm(Sepal.Width ~ Sepal.Length*Petal.Length, data = data)
summary(ols_interaction)
```

### Advanced Syntax: Non-linear regression and Interaction term 

```{r}
ols_q_i <- lm(Sepal.Width ~ Sepal.Length*as.factor(Species) + I(Sepal.Length^2), data = data)
summary(ols_q_i)
```

## Obtaining the residuals

Recall it is possible the measure the error between the model and response variables. Given least squares fit, 

$$r_i = y_i - \beta_0 +\beta_1 X_{1i} + \beta_2 X_{2i} + \cdots + \beta_n X_{ni}$$
r is the residual vector and $\sum_{i=1}^m r_i^2$ is the residual standard error. 

It is possible to obtain residuals from the output of `lm`
```{r}
res <- ols$residuals
head(res)
```

## Obtaining the fitted values

It is also possible to get predicted model values from the output of `lm`.

```{r}
pred <- ols$fitted.values
head(data.frame(pred=pred,orig=data$Sepal.Width ))
```

## Obtaining the ANOVA table


```{r}
anova(ols)
```

## Obtaining the variance-covariance matrix


```{r}
vcov(ols)
```

## Obtaining confidence intervals for the coefficients


```{r}
confint(ols, level = 0.95)
```

## Obtaining confidence interval of the mean response

```{r}
predict(ols, data.frame(Sepal.Length=4.0, Species="setosa"), interval="confidence", level = 0.95, se.fit=TRUE)
```

## Obtaining prediction limits for new observation

```{r}
predict(ols, data.frame(Sepal.Length=4.0, Species="setosa"),  interval="prediction", level = 0.95, se.fit=TRUE)
```

## Obtaining confidence band for the entire regression line

```{r}
alpha = 0.05
n = dim(iris)[1]
ci <- predict(ols, data.frame(Sepal.Length=4.0, Species="setosa"), interval="confidence", level= 1-alpha, se.fit=TRUE)
yh.hat <- ci$fit[1]
se.yh.hat <- ci$se.fit
w <- sqrt(2*qf(1-alpha, 2, n-2))
lower_bound <- yh.hat - w*se.yh.hat
upper_bound <- yh.hat + w*se.yh.hat
band <- c(lower_bound, upper_bound)
band
```

## Standardized regression

```{r}
lm.beta(ols)
```

# Model selection

## General linear test approach

```{r}
anova(ols_q_i, ols_interaction)
```

```{r}
anova(ols_q_i, ols_quadratic)
```

```{r}
anova(ols_quadratic, ols)
```

```{r}
anova(ols_interaction, ols)
```

## AIC, BIC, and Adjusted R2

```{r}
aic <- c(AIC(ols_quadratic), AIC(ols_interaction))
bic <- c(BIC(ols_quadratic), BIC(ols_interaction))
ar2 <- c(summary(ols_quadratic)$adj.r.squared, summary(ols_interaction)$adj.r.squared)
cbind(aic, bic, ar2)
```

## Step wise selection

```{r}
base <- lm(Sepal.Width ~ 1, data=data)
retailer_quadratic <- lm(Sepal.Width ~ . + I(Sepal.Length^2), data=data)
retailer_interaction <- lm(Sepal.Width ~ . + Sepal.Length*Species, data=data)
step_1 <- step(base, scope = list(upper=retailer_quadratic, lower= ~1), direction = "both", trace=TRUE)
step_2 <- step(base, scope = list(upper=retailer_interaction, lower= ~1), direction = "both", trace=TRUE)
```

```{r}
summary(step_1)
summary(step_2)
```

# Diagnostics

## Test for lack of fit 

```{r}
full <- lm(Sepal.Width ~ as.factor(Sepal.Length)*as.factor(I(Sepal.Length^2))*as.factor(Species), data=data)
anova(ols_quadratic, full)
```
## Normal probability plot

```{r}
std_res <- rstandard(ols_quadratic)
qqnorm(std_res, 
     ylab="Standardized Residuals", 
     xlab="Normal Scores") 
qqline(std_res)
```
## Residual plots

```{r}
par(mfrow=c(2,2))
plot(res ~ pred, xlab="Fitted", ylab="Residual", main="Residual plot against fitted values")
abline(h=0)
plot(res ~ data$Sepal.Length, xlab="X", ylab="Residual", main="Residual plot against X")
abline(h=0)
```
## Test for multicollinearity

 See also function vif() in package "car"
```{r}
VIF <- rep(0,2)
VIF[1] <- 1/(1-summary(lm(Sepal.Width ~ Sepal.Length, data = data))$r.squared)
VIF[2] <- 1/(1-summary(lm(Sepal.Width ~ as.factor(Species), data = data))$r.squared)
VIF
```
## Test for heteroskedasticity


```{r}
bptest(ols_quadratic, studentize = FALSE)
```
## Test for outlying Y observations--studentized deleted residual

```{r}
p <- 5 # numer of parameters
case <- c(1:n) # n defined above
plot(case, rstudent(ols_quadratic), type="l", xlab="Case Numbers", 
     ylab="Studentized Deleted Residuals", main="Test for Outlying Y Values")
text(case, rstudent(ols_quadratic), case)
alpha <- 0.05
crit <- qt(1-alpha/2/n, n-p-1)
which(abs(rstudent(ols_quadratic)) >=crit ) # Here there's no evidence of outlying Y observations
```
## Test for outlying X observations--hat matrix leverage values

```{r}
leverage <- hatvalues(ols_quadratic)
plot(case, leverage, type="l", xlab="Case Numbers", 
     ylab="Leverage", main="Test for Outlying X Values")
text(case, leverage, case)
abline(h=0.5, col=2)
X_out <- which(leverage>0.5)
leverage[X_out] # Here there's no outlying X observations
```
## Tests for influential observations
### Cook's distance
```{r}
plot(case, cooks.distance(ols_quadratic), type="l", xlab="Case Numbers", 
     ylab="Cook's Distance", main = "Test for Influential Values: Cook's Distance")
text(case, cooks.distance(ols_quadratic))

inf_obs <- which(cooks.distance(ols_quadratic)>0.5)
inf_obs
```
### DFFITS

```{r}
plot(case, dffits(ols_quadratic), type="l", xlab="Case Numbers", 
     ylab="DFFITS", main = "Test for Influential Values: DFFITS")
text(case, dffits(ols_quadratic))

inf_obs2 <- which(abs(dffits(ols_quadratic))>2/sqrt(p/n))
inf_obs2
```

### DFBETAS
```{r}
inf_obs3 <- which(abs(dfbeta(ols_quadratic))>2/sqrt(n))
inf_obs3
```